---
title: "DATASCI 306, Homework 10b, Winter 2026"
author: "James Souilliere"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Background: Six Degrees of Kevin Bacon

The "Six Degrees of Kevin Bacon" is a parlor game based on the idea that any
actor in Hollywood can be connected to Kevin Bacon through a chain of shared
movie appearances in at most six steps. For example, if Actor A appeared in a
movie with Actor B, and Actor B appeared in a movie with Kevin Bacon, then Actor
A has a "Bacon number" of 2.

In this homework, you will build an interactive Shiny app that implements this
game using the same `movie_actors.csv` dataset from Homework 10a.

### Data

```{r}
movie_actors <- read_csv("movie_actors.csv")
glimpse(movie_actors)
```

---


\newpage

## Question 1 (10 points): Six Degrees of Kevin Bacon App

In this question you will complete an interactive version of the Six Degrees of
Kevin Bacon game as a Shiny app. The player navigates from a starting actor
toward Kevin Bacon by choosing movies and co-stars, one step at a time.

A starter file `bacon_app.R` is provided with the UI, data loading, and helper
functions already written. **Your job is to fill in the server logic** --- the
sections marked `# TODO` in the file. Do not modify the UI or the helper
functions.

### Game Flow

1. **Start:** The user clicks "New Game". The app picks 5 random actors and
   displays them as radio button choices.
2. **Pick an actor:** The user selects an actor and clicks "Select". The app
   shows that actor's movies as the new choices.
3. **Pick a movie:** The user selects a movie and clicks "Select". The app shows
   the cast of that movie (excluding already-visited actors) as choices.
4. **Repeat** steps 2--3 until Kevin Bacon is selected (win!) or the player has
   picked 6 actors without finding him (lose).

### What to complete

Open `bacon_app.R` and find the four `# TODO` blocks inside the `server`
function. Read the comments above each block carefully --- they describe exactly
what that section should do. Here is a summary:

1. **TODO 1 --- `observeEvent(input$new_game, ...)`:** Reset the game state and
   sample 5 random starting actors, weighted by number of movies (pass
   `prob = actor_weights` to `sample()` so recognizable actors appear more
   often). Use `updateRadioButtons()` to set the choices for the `"selection"`
   input.

2. **TODO 2 --- `observeEvent(input$choose, ...)`:** This is the main game logic.
   When the user clicks "Select":
   - If in the `"actor"` phase: record the chosen actor in the path and visited
     list, increment the step count, check for win/loss, then look up the
     actor's movies (filtering out any already in `game$visited_movies` with
     `setdiff()`) and switch to the `"movie"` phase. The provided `get_movies()`
     helper returns movies sorted by release year, so options appear
     chronologically.
   - If in the `"movie"` phase: record the chosen movie in the path **and in
     `game$visited_movies`**, look up the cast (excluding visited actors, but
     always including Kevin Bacon if present), and switch back to the
     `"actor"` phase.

3. **TODO 3 --- `output$path_display`:** Use `renderUI` to display the path taken
   so far. The `game$path` vector alternates between actor names and movie
   titles. Display actors in **bold** and movies in *italics*, separated by
   arrows.

4. **TODO 4 --- `output$result_message`:** Use `renderText` to show a win message
   (with Bacon number), a loss message, or nothing depending on the game state.

### Provided helpers

These are already defined in the starter file --- use them in your TODOs:

- `get_movies(actor_name)` --- returns a character vector of movies for an
  actor, sorted by release year (ascending)
- `get_cast(movie_name)` --- returns a character vector of actors in a movie
- `non_bacon_actors` --- character vector of all actors except Kevin Bacon
- `actor_weights` --- named numeric vector of movie counts for each actor in
  `non_bacon_actors` (use as `prob =` when sampling starting actors)

### Hints

- To dynamically change the radio button choices from the server, use
  `updateRadioButtons(session, "selection", choices = new_choices_vector)`.
  The `session` object is available as a parameter of the server function.

- `game$phase` is either `"actor"` or `"movie"` --- check it to decide what to
  do when `input$choose` is clicked.
- `game$visited` tracks actors already chosen so they can be excluded from
  future cast lists. Use `setdiff()` to remove them.
- `game$visited_movies` plays the analogous role for movies: update it in the
  movie phase, and `setdiff()` it from `get_movies()` in the actor phase so
  the player can't revisit a movie already on the path.
- The step counter (`game$step`) counts *actor* selections only, not movie
  selections. The game ends at 6 actor selections.
- Use `req(game$active, input$selection)` at the top of your `observeEvent` for
  the choose button so it only fires when a game is in progress.
- For the path display, `seq_along(game$path)` with odd indices = actors and
  even indices = movies.

End the file with `shinyApp(ui, server)` (already provided). Test with
`shiny::runApp("bacon_app.R")`. Your completed code will be included
automatically:

```{r, code=readLines("bacon_app.R"), eval=FALSE}
```
